SNP Analysis

### 2018-07-18 SNP_Prediction_Analysis.Rmd
...

adjust.cols<-function(mtx, means=NULL, sds=NULL) {
  if (!is.null(means)) {
    if (ncol(mtx) != length(means))
      stop("Vector of means needs to match the columns in the matrix")
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i], means[i], sds[i])
  } else {
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i])
  }
  return(mtx)
}

@@

nm.rows = grep('BLD|GASTRIC', rownames(mergedSegs))
normalSegs = mergedSegs[nm.rows,]
normalArms = mergedArms[nm.rows,]

nrow(normalSegs) == nrow(normalArms)

mergedSegs = mergedSegs[-nm.rows,]
mergedArms = mergedArms[-nm.rows,]

nrow(mergedSegs) == nrow(mergedArms)

@@

grid.arrange(p1,p2)
```

@@

```{r normals, message=F, fig.height=10, fig.width=8}
nm = melt(normalSegs)
be = melt(mergedSegs[-nm.rows,])

lims = range(nm$value, be$value)

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric',subtitle=paste('variance=',var(nm$value),sep=''),x='') + 
    theme(axis.text.x = element_blank()) + ylim(lims),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples',subtitle=paste('variance=',var(be$value),sep=''),x='') + 
    theme(axis.text.x = element_blank()) + ylim(lims)
)

@@

lims = range(segs$medLRR, normalSegs[randNormal,])
grid.arrange(
  ggplot(segs) + facet_grid(~chr, space='free_x', scales='free') + ylim(lims) +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='ASCAT median LRR', subtitle=pt),
  ggplot(melt(as.matrix(normalSegs[randNormal,])), aes(Var1, value)) + geom_point() + ylim(lims) +
    labs(title='CN Adjusted & winsorized') + theme(axis.text.x=element_blank()),
top='Random BLD or gastric normal', bottom=randNormal)

@@

lims = range(segs$medLRR, mergedSegs[sub('\\.LogR','',randBE),])
grid.arrange(
  ggplot(segs) + facet_grid(~chr, space='free_x', scales='free')  +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='Median LRR, ASCAT segments', subtitle=pt), 
  ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randBE),])), aes(Var1, value)) + geom_point() +
    theme(axis.text.x=element_blank()) + labs(title='Winsorized LRR, binned'),  
  ggplot(melt(adjust.cols(mergedSegs[sub('\\.LogR','',randBE),,drop=F], z.mean, z.sd)), aes(Var2, value)) + geom_point() +
    theme(axis.text.x=element_blank()) + labs(title='Col adjusted LRR, binned'),  
top='BE', bottom=randBE)

@@

Using the mean bin values (merged during tiling).
```{r, warning=F, message=F, fig.width=10, fig.height=8}
# z.mean[is.na(z.mean)] = median(z.mean,na.rm=T)
# z.arms.mean[is.na(z.arms.mean)] = median(z.arms.mean,na.rm=T)
# 
# z.sd[is.na(z.sd)] = median(z.sd,na.rm=T)
# z.arms.sd[is.na(z.arms.sd)] = median(z.arms.sd,na.rm=T)

ns = adjust.cols(normalSegs, z.mean, z.sd)
#ns = apply(normalSegs, 2, unit.var)
range(ns)

na = adjust.cols(normalArms, z.arms.mean, z.arms.sd)
#na = apply(normalArms, 2, unit.var)
range(na)

ms = adjust.cols(mergedSegs, z.mean, z.sd)
#ms = apply(mergedSegs, 2, unit.var)
range(ms)
dim(ms)

ma = adjust.cols(mergedArms, z.arms.mean, z.arms.sd)
#ma = apply(mergedArms, 2, unit.var)
range(ma)
dim(ma)

nm = melt(ns)
be = melt(ms)

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric') + 
    theme(axis.text.x=element_blank()) +ylim(range(nm$value, be$value)),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples') + 
    theme(axis.text.x=element_blank()) + ylim(range(nm$value, be$value)), top='Unit scaled')

grid.arrange(
  ggplot(melt(ms[sub('\\.LogR','',randBE),,drop=F]), aes(Var2, value)) + geom_point() +
    theme(axis.text.x=element_blank()) + labs(title='Col adjusted LRR, binned'),  
  ggplot(melt(subtract.arms(ms[sub('\\.LogR','',randBE),,drop=F],ma[sub('\\.LogR','',randBE),,drop=F])), aes(Var2, value)) +
    geom_point() + theme(axis.text.x=element_blank()) + labs(title='Arms subtracted col adjusted LRR, binned'),  
  top=randBE)

@@

```{r echo=F,message=F, fig.width=10, fig.height=8}
cxN = score.cx(ns, 1)
arrayDfNorm = subtract.arms(ns, na)
#arrayDfNorm = cbind(ns, na)
arrayDfNorm = cbind(arrayDfNorm, 'cx'=unit.var(cxN, mn.cx, sd.cx))

probN = predict(fitV, newx=arrayDfNorm, s=l, type='response')
length(which(probN > 0.5))/nrow(probN) # FPR
length(which(probN <= 0.5))/nrow(probN)

preds.normal = cbind.data.frame(rownames(probN),probN, predict(fitV, newx=arrayDfNorm, s=l, type='link'))
colnames(preds.normal) = c('Samplename','Prob', 'RR')

cx = score.cx(ms, 1)
#arrayDf = cbind(ms,ma)
arrayDf = subtract.arms(ms, ma)

arrayDf[,589:ncol(arrayDf)] = arrayDf[,589:ncol(arrayDf)]+abs(mean(arrayDf[,589:ncol(arrayDf)]))

arrayDf = cbind(arrayDf, 'cx'=unit.var(cx, mn.cx, sd.cx))

grid.arrange(
  ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + 
    geom_point(data=melt(arrayDf[,589:634]), col='red') + labs(title='SNP'),
  ggplot(melt(allDf), aes(Var2, value)) + geom_point() +
      geom_point(data=melt(allDf[,589:634]), col='red') + labs(title='sWGS'),
nrow=1)

grid.arrange(
  ggplot( melt(allDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS'),
  ggplot( melt(arrayDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP arrays'),
  ncol=1
)

@@

```{r, warning=F, fig.height=8}
#offset = log(0.0225)
# preds = preds %>% dplyr::mutate( 
#   'Adj. RR'=RR+offset,  
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

@@

length(which(preds.normal$Prob > 0.5))/nrow(preds.normal)

qcdata.be = subset(qcdata, SampleType == 'BE')

get.qcinfo<-function(vec) {
  endo = vec[2]
  pt = vec[1]
  y = subset(qcdata.be, PatientID == pt & EndoID == endo)
  y = y %>% group_by(Status, Timepoint) %>%  dplyr::summarise( mean(Ploidy), mean(Purity), mean(Goodnessoffit), mean(SCA.Ratio) )
  cbind.data.frame('PatientID'=pt, 'EndoID'=endo, y, stringsAsFactors = F)
}

qc = do.call(rbind.data.frame, c( lapply(preds$Samplename, function(x) get.qcinfo( unlist(strsplit(x, '_'))) ), stringsAsFactors=F))

pred.qc = cbind.data.frame(preds, qc)
#dim(pred.qc)
#table(pred.qc$Status)

@@

@@

roc = pROC::roc(Status ~ Risk, data=subset(pred.qc, !exclude), auc=T, ci=T, of='thresholds')
roc.plot(roc) + labs(title='Excluded low sca/NP tetra), Risk')

@@

rocpp = pROC::roc(Status ~ max.Risk, data=per_pt, auc=T, ci=T, of='thresholds')
roc.plot(rocpp) + labs(title='Max per patient (excluded low sca/NP tetra), Risk')

@@

rocppex = pROC::roc(Status ~ max.Risk, data=per_pt_ex, auc=T, ci=T, of='thresholds')
roc.plot(rocppex) + labs(title='Max per patient, Risk')

@@

roct2 = pROC::roc(Status ~ Risk, data=subset(pred.qc, Timepoint != 'T1' & !exclude), auc=T, ci=T, of='thresholds')
roc.plot(roct2) + labs(title='T2, Risk')

@@

@@

# Quick check
nrow(subset(ex.pred, Risk == 'High' & Status == 'P'))/table(ex.pred$Status)['P']
nrow(subset(ex.pred, Risk == 'Low' & Status == 'P'))/table(ex.pred$Status)['P'] #False negatives

ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + 
  geom_point(data=melt(arrayDf['1006_190V',,drop=F]), color='red') + 
  geom_point(data=melt(arrayDf['1044_39Z',,drop=F]), color='blue')

ggplot(melt(dysplasia.df), aes(Var2, value)) + geom_point()

grid.arrange(
  ggplot(melt(dysplasia.df[,589:ncol(dysplasia.df)]), aes(Var2, value)) + geom_point(),
  ggplot(melt(arrayDf[,589:ncol(arrayDf)]), aes(Var2, value)) + geom_point())

nrow(subset(ex.pred, Risk == 'Low' & Status == 'NP'))/table(ex.pred$Status)['NP']
nrow(subset(ex.pred, Risk == 'High' & Status == 'NP'))/table(ex.pred$Status)['NP'] # False positive

>>>>

adjust.cols<-function(mtx, means=NULL, sds=NULL, na.replace=0) {
  if (!is.null(means)) {
    if (ncol(mtx) != length(means))
      stop("Vector of means needs to match the columns in the matrix")
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i], means[i], sds[i])
  } else {
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i])
  }
  mtx[is.na(mtx)] = na.replace

  return(mtx)
}

@@

# Adjust for ploidy
for (i in 1:nrow(mergedSegs)) {
  sample = rownames(mergedSegs)[i]
  x = unlist(strsplit(sample, '_'))
  ploidy = mean(subset(qcdata, SampleType == 'BE' & PatientID == x[1] & EndoID == x[2])$Ploidy)
  if (grepl('BLD|gastric', sample, ignore.case = T)) ploidy = 2
  mergedSegs[i,] = mergedSegs[i,]-(ploidy-1)
  mergedArms[i,] = mergedArms[i,]-(ploidy-1)
}

nm.rows = grep('BLD|GASTRIC', rownames(mergedSegs))
#normalSegs = mergedSegs[nm.rows,]
#normalArms = mergedArms[nm.rows,]

# nrow(normalSegs) == nrow(normalArms)
# 
# mergedSegs = mergedSegs[-nm.rows,]
# mergedArms = mergedArms[-nm.rows,]
# 
# nrow(mergedSegs) == nrow(mergedArms)

@@

grid.arrange(p1,p2,ncol=1)
```

@@

```{r normals, message=F, fig.height=10, fig.width=8}
#nm = melt(normalSegs)
nm = melt(mergedSegs[nm.rows,])
be = melt(mergedSegs[-nm.rows,])

lims = range(nm$value, be$value, na.rm=T)

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric',subtitle=paste('variance=',var(nm$value),sep=''),x='') + 
    theme(axis.text.x = element_blank()) + ylim(lims),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples',subtitle=paste('variance=',var(be$value,na.rm=T),sep=''),x='') + 
    theme(axis.text.x = element_blank()) + ylim(lims)
)

@@

lims = range(segs$medLRR, normalSegs[randNormal,])
# grid.arrange(
#   ggplot(segs) + facet_grid(~chr, space='free_x', scales='free') + ylim(lims) +
#     geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
#     theme(axis.text.x=element_blank() ) + labs(title='ASCAT median LRR', subtitle=pt),
#   ggplot(melt(as.matrix(normalSegs[randNormal,])), aes(Var1, value)) + geom_point() + ylim(lims) +
#     labs(title='CN Adjusted & winsorized') + theme(axis.text.x=element_blank()),
# top='Random BLD or gastric normal', bottom=randNormal)

@@

#lims = range(segs$medLRR, mergedSegs[sub('\\.LogR','',randBE),])
# grid.arrange(
#   ggplot(segs) + facet_grid(~chr, space='free_x', scales='free')  +
#     geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
#     theme(axis.text.x=element_blank() ) + labs(title='Median LRR, ASCAT segments', subtitle=pt), 
#   ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randBE),])), aes(Var1, value)) + geom_point() +
#     theme(axis.text.x=element_blank()) + labs(title='Winsorized LRR, binned'),  
#  ggplot(melt(adjust.cols(mergedSegs[sub('\\.LogR','',randBE),,drop=F], z.mean, z.sd)), aes(Var2, value)) + geom_point() + theme(axis.text.x=element_blank()) + labs(title='Col adjusted LRR, binned'),  
#top='BE', bottom=randBE)

@@

Using the mean bin values (merged during tiling).
```{r, warning=F, message=F, fig.width=10, fig.height=8}
#ns = adjust.cols(normalSegs, z.mean, z.sd)
#ns = adjust.cols(normalSegs)
#range(ns)

#na = adjust.cols(normalArms, z.arms.mean, z.arms.sd)
#na = adjust.cols(normalArms)
#range(na)

#ms = adjust.cols(mergedSegs, z.mean, z.sd)
ms = adjust.cols(mergedSegs)
range(ms)
dim(ms)

# mergedSegs['997_122V',which(mergedSegs['997_122V',] <= 0)]
# ms['997_122V',which(mergedSegs['997_122V',] <= 0)]

#ma = adjust.cols(mergedArms, z.arms.mean, z.arms.sd)
ma = adjust.cols(mergedArms)
range(ma)
dim(ma)

nm = melt(ms[nm.rows,])
be = melt(ms[-nm.rows,])

lims = range(nm$value, be$value, na.rm = T)
grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2, color=Var1)) + geom_point() + labs(title='Blood/gastric') + 
    theme(axis.text.x=element_blank(),legend.position='none') +ylim(lims),
  ggplot(be, aes(Var2, value, group=Var2, color=Var1)) + geom_point() + labs(title='BE samples') + 
    theme(axis.text.x=element_blank(),legend.position='none') + ylim(lims), top='Unit scaled')

@@

```{r echo=F,message=F, fig.width=10, fig.height=8}
#cxN = score.cx(ns, 1)
#arrayDfNorm = subtract.arms(ns, na)
#arrayDfNorm = cbind(ns, na)
#arrayDfNorm = cbind(arrayDfNorm, 'cx'=unit.var(cxN, mn.cx, sd.cx))

#probN = predict(fitV, newx=arrayDfNorm, s=l, type='response')
#length(which(probN > 0.5))/nrow(probN) # FPR
#length(which(probN <= 0.5))/nrow(probN)

#preds.normal = cbind.data.frame(rownames(probN),probN, predict(fitV, newx=arrayDfNorm, s=l, type='link'))
#colnames(preds.normal) = c('Samplename','Prob', 'RR')

cx = score.cx(ms, 1)
#arrayDf = cbind(ms,ma)
arrayDf = subtract.arms(ms, ma)
#arrayDf = cbind(arrayDf, 'cx'=unit.var(cx, mn.cx, sd.cx))
arrayDf = cbind(arrayDf, 'cx'=unit.var(cx))

lims=range(arrayDf, allDf,na.rm=T)
grid.arrange(
  ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + ylim(lims) +
    geom_point(data=melt(arrayDf[,589:634]), col='red') + labs(title='SNP'),
  ggplot(melt(allDf), aes(Var2, value)) + geom_point() + ylim(lims) +
      geom_point(data=melt(allDf[,589:634]), col='red') + labs(title='sWGS'),
nrow=1)

```{r, echo=F, fig.height=12, fig.width=6}
grid.arrange(
  ggplot( melt(allDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS'),
  ggplot( melt(arrayDf), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP arrays'),
  ncol=1
)

grid.arrange(
  ggplot(melt(allDf[,1,drop=F]), aes(sample=value)) + stat_qq(),
  ggplot(melt(arrayDf[,1,drop=F]), aes(sample=value)) + stat_qq(),
  ncol=1)

for (i in 1:ncol(arrayDf)) {
  arrayDf[,i] = unit.var(arrayDf[,i], mean(allDf[,i]), sd(allDf[,i]))
}

qqnorm()
sd(allDf[,1])

mean(arrayDf[,1])
sd(arrayDf[,1])

@@

```{r, warning=F, fig.height=8}

pnr = grep('BLD|gastric', rownames(preds), ignore.case = T)
preds.normal = preds[pnr,]

length(which(preds.normal$Prob > 0.5))/nrow(preds.normal) # FPR
length(which(preds.normal$Prob < 0.5))/nrow(preds.normal)

#offset = log(0.0225)
# preds = preds %>% dplyr::mutate( 
#   'Adj. RR'=RR+offset,  
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

@@

qcdata.be = subset(qcdata, SampleType == 'BE')
get.qcinfo<-function(vec) {
  endo = vec[2]
  pt = vec[1]
  y = subset(qcdata.be, PatientID == pt & EndoID == endo)
  y = y %>% group_by(Status, Timepoint) %>%  dplyr::summarise( mean(Ploidy), mean(Purity), mean(Goodnessoffit), mean(SCA.Ratio) )
  cbind.data.frame('PatientID'=pt, 'EndoID'=endo, y, stringsAsFactors = F)
}

qc = do.call(rbind.data.frame, c( lapply(preds$Samplename[-pnr], function(x) get.qcinfo( unlist(strsplit(x, '_'))) ), stringsAsFactors=F))

pred.qc = cbind.data.frame(preds[-pnr,], qc)
#dim(pred.qc)
#table(pred.qc$Status)

@@

## First of all we have samples from patients that probably had a genome duplication, there's no way to measure that in our sWGS data
#subset(pred.qc, `mean(Ploidy)` > 2.7)

```{r echo=F}

@@

@@

@@

@@

@@

subset(pred.qc, `mean(SCA.Ratio)` < 0.1 & Status == 'P' & !exclude & Prob < 0.5)

@@

# Quick check
nrow(subset(ex.pred, Prob > 0.5 & Status == 'P'))/table(ex.pred$Status)['P']
nrow(subset(ex.pred, Prob < 0.5 & Status == 'P'))/table(ex.pred$Status)['P'] #False negatives

nrow(subset(ex.pred, Prob > 0.5 & Status == 'NP'))/table(ex.pred$Status)['NP'] # False pos
nrow(subset(ex.pred, Prob < 0.5 & Status == 'NP'))/table(ex.pred$Status)['NP'] #

# ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + 
#   geom_point(data=melt(arrayDf['1006_190V',,drop=F]), color='red') + 
#   geom_point(data=melt(arrayDf['1044_39Z',,drop=F]), color='blue')
